---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


##### The data for this analysis comes from Roger Peng's github account https://github.com/rdpeng/RepData_PeerAssessment1 and was forked on 6/6/2015.

```{r echo=TRUE}
proj_data <- read.csv("./activity.csv")
head(proj_data)
```

### The purpose of this analysis is to determine and illustrate the impact of missing values on this dataset. Further, we examine the differences between weekends and weekdays.
### First, two analyses are performed - first with missing data ignored and second with missing values replaced with average values
### First Analysis
#### Ignoring missing values, we examine the mean and median number of steps per day
```{r echo=TRUE}
daily_steps <- tapply(proj_data$steps, proj_data$date, sum, na.rm = TRUE)
hist(daily_steps, xlab = "Steps per Day", ylab = "# of Days", main = "Histogram of Steps per Day", breaks = 25)
```

#### The calculated mean and median number of steps per day are:

```{r echo = TRUE}
cat("Mean = ", mean(daily_steps),"\nMedian = ", median(daily_steps))
```

#### Omitting missing values, we view the average daily activity pattern using a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days
```{r echo=TRUE}
mean_interval_steps <- tapply(proj_data$steps, proj_data$interval, mean, na.rm = TRUE)
plot(names(mean_interval_steps),
     mean_interval_steps,  
     type="l", 
     xlab = "Time of Day in 5 Min Increments", 
     ylab = "Mean Steps Taken in Interval")
```

#### The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps is:
```{r echo=TRUE}
names(which.max(mean_interval_steps))
```
and the value is
```{r echo = TRUE}
max(mean_interval_steps)
```

### Now we impute values for the NA data in the dataset and once again run a second analysis to see if this significantly changes our view of the results
#### The number of observations (rows) with missing values is:
```{r echo = TRUE}
sum(is.na(proj_data$steps))
```
#### Replace the missing values with the mean for that period across all days and store in a new dataframe
```{r echo = TRUE}
nas <- is.na(proj_data$steps)
proj_data_no_nas <- proj_data
proj_data_no_nas$steps[nas] <- mean_interval_steps[as.factor(proj_data_no_nas$interval[nas])]
```

#### The histogram of steps per day shifts with the imputed missing values and the maximum number of steps in a day increases

```{r echo=TRUE}
daily_steps_no_nas <- tapply(proj_data_no_nas$steps, proj_data_no_nas$date, sum)
hist(daily_steps_no_nas, xlab = "Steps per Day", ylab = "# of Days", main = "Histogram of Steps per Day", breaks = 25)
```

#### The calculated mean and median number of steps increases slightly because we replaced the missing values with the mean for the period across all days.

```{r echo = TRUE}
cat("Mean = ", mean(daily_steps_no_nas),"\nMedian = ", median(daily_steps_no_nas))
```

#### Since we replaced the missing values with the mean, we would expect the plot of the mean steps in each period to remain unchanged, which is exactly what we see

```{r echo=TRUE}
mean_interval_steps_no_nas <- tapply(proj_data_no_nas$steps, proj_data_no_nas$interval, mean)
plot(names(mean_interval_steps_no_nas),
     mean_interval_steps_no_nas,  
     type="l", 
     xlab = "Time of Day in 5 Min Increments", 
     ylab = "Mean Steps Taken in Interval")
```

#### The 5-minute interval, on average across all the days in the dataset, that contains the maximum number of steps remains unchanged:
```{r echo=TRUE}
names(which.max(mean_interval_steps_no_nas))
```
and the value is
```{r echo = TRUE}
max(mean_interval_steps_no_nas)
```
### The conclusion is that while the maximum steps in a day increases with the imputed values, the impact on the mean and median is not significant
## Now we examing the difference between weekends and weekdays in the number of steps taken
### Subset the data, then plot with Lattice
```{r echo = TRUE}
library(reshape2)
library(lattice)
proj_data_no_nas$day <- weekdays(as.Date(proj_data_no_nas$date,"%Y-%m-%d"),
                                 abbreviate = TRUE)
proj_data_no_nas$day[proj_data_no_nas$day %in% c("Mon","Tue","Wed","Thu","Fri")] <- "Weekday"
proj_data_no_nas$day[proj_data_no_nas$day %in% c("Sat","Sun")] <- "Weekend"
proj_data_no_nas$day <- as.factor(proj_data_no_nas$day)
mean_steps_by_day <- dcast(proj_data_no_nas, interval ~ day, value.var = "steps", fun.aggregate = mean)
mean_steps_by_dow <- melt(mean_steps_by_day, id.vars = "interval")
xyplot(mean_steps_by_dow$value ~ mean_steps_by_dow$interval|mean_steps_by_dow$variable,
       type="l",
       xlab = "Time of Day",
       ylab = "Mean Steps Taken")
```

